Modelling BCG vaccination in the UK

### 2019-06-19 rmds/08-model-development.Rmd
compile check

In the previous chapter (Chapter \@ref(direct-eff)) I estimated the direct impact of the change in BCG policy on the subset of the population who were directly impacted by the change. The time horizon of any estimated impact was limited by the available data. If there is a non-negligible amount of Tuberculosis (TB) transmission amongst the UK born then any change in BCG vaccination policy will also have in-direct impacts via on-wards transmission. In order to capture these impacts a dynamic transmission model (see Chapter \@ref(introduction)) may be used, this explicitly models the rate that individuals are infected using the mass action assumption. A dynamic transmission model also allows estimates to be made of the long term impact of BCG policy changes via model simulation, although this does require some assumptions to be made about future population demographics and TB incidence rates in the non-UK born population.

This chapter presents the development and parameterisation of a, semi-stochastic, dynamic model of TB transmission, incorporating BCG vaccination, in the UK born population of England. The key features of TB transmission, and BCG vaccination are discussed, with details of pertinent TB models given. An appropriate model structure for answering the study question is then outlined, along with a justification of the choices made and details of required sensitivity analyses. The model structure is then defined mathematically and parameterised using literature sources as well as data from the Enhanced Tuberculosis Surveillance System (ETS), Labour Force Survey (LFS) and Office for National Statistics (ONS) (see Chapter \@ref(data)). The assumptions made during model building and parameterisation are highlighted in preparation for evaluation during model fitting.

@@

When developing an infectious disease dynamic model there is a trade-off between reproducing reality and interpretability.[@Anderson1991] A model that includes all known features of a disease may not be able to answer questions of interest as it is too complex to interpret or because data does not exist to calibrate many of its parameters. A highly complex model, or indeed an overly simplistic one, may also be at risk of bias. The optimal model is therefore as parsimonious as possible, whilst still capturing the key features of a disease and making best use of all available data.[@Anderson1991] In this section the key features of TB and BCG vaccination that must be captured in order to produce meaningful output are discussed, as well as the features that can be excluded for this study question. Data from the Enhanced Tuberculosis Surveillance (ETS) (Chapter \@ref(data)) is used to support evidence from the literature. Further background information can be found in Chapter \@ref(background) and Chapter \@ref(data).

@@

1. **Latency** - after an initial infection 5-10% of individuals develop symptomatic TB within 1-2 years. The majority of individuals enter a latent state in which they passively carry TB mycobacterium but are not symptomatic. Reactivation of the bacilli can then occur many years later due to a lose of immune control.[@Gideon2011a] Simplistically latent TB may be modeled with a single latent compartment[@Brooks-Pollock2010a], more commonly an additional transition rate between the susceptible and active disease states is added.[@Menzies2018] This represents rapid progression to active disease, and slower progression via a low risk latent stage. Both of these model structures have been shown to not fit activation data well.[@Menzies2018; @Ragonnet2017] More complex structures that are commonly used incorporate either parallel or serial latency (Figure \@ref(latency-flow-diag)). Both of these structures incorporate both slow and fast latent periods and have been shown to produce identical activation dynamics.[@Ragonnet2017] This is unfortunate as they represent two disparate biological mechanisms, with the serial assumption representing decreasing risk over time for individuals and the parallel assumption suggesting that a subset of individuals are at a greater risk of developing active TB disease. For models that seek to investigate interventions targeted at latent cases this structural uncertainty is problematic. However, as BCG vaccination occurs prior to infection both structures will produce comparable results for study questions evaluating this intervention. The model presented here uses a serial latent structure. This is commonly used in the literature; simplifies modelling other aspects of TB; and has a plausible biological underpinning.[@Ragonnet2017]

@@

1. **Smear status** - microscopic analysis of sputum smear samples for acid-fast bacilli is widely used as a means of diagnosis for TB. There is evidence that smear positive cases are responsible for the majority of transmission,[@PMID:13148535] with smear negative cases contributing approximately 76% (95% CI 70%, 80%) less to transmission than smear positive cases.[@Tostmann2008] The proportion of smear positive cases varies with age,[@Piccini2014] with 30.2% (95% CI 26.3%, 33.7%) in 0-14 year-olds, 65.2% (95% CI 64.2%, 66.2%) in 15-59 year-olds and 53.6% (95% CI 51.9%, 55.3%) in 60-89 year-olds in the ETS between 2000 and 2015. the model presented here includes sputum status via the force of infection.

@@

1. **Non-UK born TB Cases** - TB incidence in England is highly heterogeneous with over 70% of cases occurring in the non-UK born population.[@PHE2017] The age distribution of cases in the UK born and non-UK born populations differ, with the UK born population having a relatively uniform distribution. Meanwhile, the non-UK born have higher incidence rates in those aged 80 years and older (69.3 per 100,000 in 2016), those aged 75 to 79 years (62.9 per 100,000 in 2016) and those aged 25-29 years old (61.6 per 100,000 in 2016).[@PHE2017] Exposure to England's BCG vaccination policy is difficult to assess for the non-UK born as is the degree of transmission occurring in the UK as opposed to cases being imported from abroad, or acquired from trips to cases countries of origin. For this reason the model presented here does not explicitly include non-UK born cases. Instead it imports non-UK born cases into the force of infection, with the addition of a mixing parameter that controls the degree of contact between non-UK born cases and those born in the UK.

@@

1. **Duration of protection** - BCG protection wanes with time, with the greatest protection shortly after vaccination. There is good evidence to suggest that the effectiveness of BCG vaccination lasts up to 15 years,[@Abubakar2013] and a recent study suggests that this protection may last later into adulthood in the UK born.[@Mangtani2017] However, the it's effectiveness reduces with time.

@@

The model includes the following compartments: Susceptible ($S$), high risk latent ($H$), low risk latent ($L$), active TB cases with pulmonary TB ($P$), active TB cases with extra-pulmonary TB disease only ($E$), pulmonary cases on treatment ($T_P$), and extra-pulmonary case on treatment ($T_E$). Cases that were previously infected and considered at low risk of developing active disease may be reinfected, although their latent infection provides partial protection. Treatment is assumed to be the only pathway for recovery for active TB disease, with a single rate used to model the heterogeneity of treatment times. A fraction of those on treatment are assumed to be lost to follow up, with these cases returned to active pulmonary or extra-pulmonary disease. Cases that start treatment immediately stop being infectious and upon treatment completion are treated as if they have low risk latent TB disease. TB mortality is included for both active TB cases on, and off, treatment. TB mortality is stratified by disease type and age. TB transmission is assumed to act under the mass action assumption. Non-UK born cases are included into the force of infection.

@@

The vaccination model is nested into the demographic process model and therefore vaccination is possible upon entry to each modeled age group. The target age group can be varied to represent changing BCG vaccination policy. The vaccinated population is then modeled explicitly throughout all disease compartments. The primary action of the BCG vaccine is to prevent the transition from latent to active disease, this is included for both high and low risk latent cases. Waning vaccination effectiveness has been included by stratifying vaccine effectiveness by age group. The partial protection offered by BCG vaccination against initial infection has been included as a negative modifier on the protection from latent to active disease and as a modifier on the proportion of cases that are initially infected. This allows estimates of the effectiveness of BCG vaccination at preventing active TB disease in the susceptible population to be used, as these estimates have the most robust data sources. It is assumed that latently infected individuals do not gain additional protection from re-infection from the BCG vaccine. The BCG vaccine has been modeled as being partially protective for all individuals rather than as a "take" vaccine (i.e all or nothing protection). This assumption simplifies the model and will not impact the dynamics of TB transmission, assuming that protected and unprotected BCG vaccinated individuals obey the mass action assumption (See Chapter \@ref(introduction) and [@Anderson1991]).

@@

\begin{equation}
\lambda^k_a = \frac{\beta_a}{N} \sum\limits_{i=1}^{A} \rho_i C_{ai}(\frac{M\iota_i}{\nu_a^P} + \sum\limits_{j = u,v}P^j_i)
  (\#eq:force-of-infection)
\end{equation}

Where $\iota_a$ is the age stratified number of non-UK born pulmonary cases notified in a given year, $P^{u,v}_a$ is the number of vaccinated, and unvaccinated, pulmonary TB cases, $\rho_a$ is the age-specific proportion of cases that are smear positive, $\nu^P_a$ is the rate of starting treatment for active pulmonary TB, $C_{ai}$ is the age-stratified contact matrix (Section \@ref(contact-matrix)), $\beta_a$ is the age-stratified transmission probability, and $M$ is the mixing rate between the UK born and non-UK born population.

@@

Parameters distributions were either estimated from the available data or assumed based on common values found in the literature. Where no comparable estimates were found in the literature a range of distributions have been considered, with the most plausible adopted as the baseline. The discussion of model parameters has been split into disease model parameters, vaccination model parameters, and demographic model parameters. The data sources used to estimate model parameters have been detailed although for the Enhanced TB Surveillance system (ETS) and the Labour Force Survey more detail is provided elsewhere (Chapter \@ref(data)).

>>>>

In the previous chapter (Chapter \@ref(direct-eff)) I estimated the direct impact of the change in BCG policy on the subset of the population who were directly impacted. Unfortunately, the time horizon of this estimate was limited by the available data. Additionallly, if there is a non-negligible amount of Tuberculosis (TB) transmission amongst the UK born then any change in BCG vaccination policy will also have in-direct impacts, via on-wards transmission, not captured in this estimate. Both of these limitations can be over come using a dynamic transmission model (see Chapter \@ref(introduction)), this explicitly models the rate that individuals are infected using the mass action assumption.[@Anderson1991; @Keeling2007] A dynamic transmission model also allows estimates to be made of the long term impact of BCG policy changes via model simulation.

This chapter presents the development and parameterisation of a, semi-stochastic, dynamic model of TB transmission, incorporating BCG vaccination, in the UK born population of England. The key features of TB transmission, and BCG vaccination, are discussed with details of pertinent TB models given. An appropriate model structure for answering the study question is then outlined, along with a justification of the choices made and details of required sensitivity analyses. The model structure is then defined mathematically and parameterised using literature sources as well as data from the Enhanced Tuberculosis Surveillance System (ETS), Labour Force Survey (LFS) and Office for National Statistics (ONS) (see Chapter \@ref(data)). The assumptions made during model building and parameterisation are highlighted in preparation for evaluation during model fitting (Chapter \@ref(model-fitting)).

@@

When developing an infectious disease dynamic model there is a trade-off between reproducing reality and interpretability.[@Anderson1991] A model that includes all known features of a disease may not be able to answer questions of interest as it is too complex to interpret or because data does not exist to calibrate many of its parameters. A highly complex model, or indeed an overly simplistic one, may also be at risk of bias. The optimal model is therefore as parsimonious as possible, whilst still capturing the key features of a disease and making best use of all available data.[@Anderson1991] In this section the key features of TB, and BCG vaccination, that must be captured in order to produce meaningful output are discussed, as well as the features that can be excluded for this study question. Data from the Enhanced Tuberculosis Surveillance (ETS) (Chapter \@ref(data)) is used to support evidence from the literature. Further background information can be found in Chapter \@ref(background) and Chapter \@ref(data).

@@

1. **Latency** - after an initial infection 5-10% of individuals develop symptomatic TB within 1-2 years. The majority of individuals enter a latent state in which they passively carry TB mycobacterium but are not symptomatic. Reactivation of the bacilli can then occur many years later due to a lose of immune control.[@Gideon2011a] Simplistically latent TB may be modeled with a single latent compartment[@Brooks-Pollock2010a], more commonly an additional transition rate between the susceptible and active disease states is added.[@Menzies2018] This represents rapid progression to active disease, and slower progression via a low risk latent stage. Both of these model structures have been shown to not fit activation data well.[@Menzies2018; @Ragonnet2017] More complex structures that are commonly used incorporate either parallel or serial latency (Figure \@ref(fig:latency-flow-diag)). Both of these structures incorporate both slow and fast latent periods and have been shown to produce identical activation dynamics.[@Ragonnet2017] This is unfortunate as they represent two disparate biological mechanisms, with the serial assumption representing decreasing risk over time for individuals and the parallel assumption suggesting that a subset of individuals are at a greater risk of developing active TB disease. For models that seek to investigate interventions targeted at latent cases this structural uncertainty is problematic. However, as BCG vaccination occurs prior to infection both structures will produce comparable results for study questions evaluating this intervention. The model presented here uses a serial latent structure. This is commonly used in the literature; simplifies modelling other aspects of TB; and has a plausible biological underpinning.[@Ragonnet2017]

@@

1. **Smear status** - microscopic analysis of sputum smear samples for acid-fast bacilli is widely used as a means of diagnosis for TB. There is evidence that smear positive cases are responsible for the majority of transmission,[@PMID:13148535] with smear negative cases contributing approximately 76% (95% CI 70%, 80%) less to transmission than smear positive cases.[@Tostmann2008] The proportion of smear positive cases varies with age,[@Piccini2014] with 30.2% (95% CI 26.3%, 33.7%) in 0-14 year-olds, 65.2% (95% CI 64.2%, 66.2%) in 15-59 year-olds and 53.6% (95% CI 51.9%, 55.3%) in 60-89 year-olds in the ETS between 2000 and 2015. the model presented here includes sputum status via the force of infection.[@Anderson1991; @Keeling2007]

@@

1. **Non-UK born TB Cases** - TB incidence in England is highly heterogeneous with over 70% of cases occurring in the non-UK born population.[@PHE2017] The age distribution of cases in the UK born and non-UK born populations differ, with the UK born population having a relatively uniform distribution. Meanwhile, the non-UK born have higher incidence rates in those aged 80 years and older (69.3 per 100,000 in 2016), those aged 75 to 79 years (62.9 per 100,000 in 2016) and those aged 25-29 years old (61.6 per 100,000 in 2016).[@PHE2017] Exposure to England's BCG vaccination policy is difficult to assess for the non-UK born as is the degree of transmission occurring in the UK as opposed to cases being imported from abroad, or acquired from trips to cases countries of origin. For this reason the model presented here does not explicitly include non-UK born cases. Instead it imports non-UK born cases into the force of infection,[@Anderson1991; @Keeling2007] with the addition of a mixing parameter that controls the degree of contact between non-UK born cases and those born in the UK.

@@

1. **Duration of protection** - BCG protection wanes with time, with the greatest protection shortly after vaccination. There is good evidence to suggest that the effectiveness of BCG vaccination lasts up to 15 years,[@Abubakar2013] and a recent study suggests that this protection may last later into adulthood in the UK born.[@Mangtani2017] However, the it's effectiveness reduces with time since vaccination.

@@

The model includes the following compartments: Susceptible ($S$), high risk latent ($H$), low risk latent ($L$), active TB cases with pulmonary TB ($P$), active TB cases with extra-pulmonary TB disease only ($E$), pulmonary cases on treatment ($T_P$), and extra-pulmonary case on treatment ($T_E$). Cases that were previously infected and considered at low risk of developing active disease may be reinfected, although their latent infection provides partial protection. Treatment is assumed to be the only pathway to recovery for active TB disease, with a single rate used to model the heterogeneity of treatment times. A fraction of those on treatment are assumed to be lost to follow up, with these cases returned to active pulmonary or extra-pulmonary disease. Cases that start treatment immediately stop being infectious and upon treatment completion are treated as if they have low risk latent TB disease. TB mortality is included for both active TB cases on, and off, treatment. TB mortality is stratified by disease type and age. TB transmission is assumed to act under the mass action assumption.[@Anderson1991; @Keeling2007] Non-UK born cases are included into the force of infection.[@Anderson1991; @Keeling2007]

@@

The vaccination model is nested into the demographic process model and therefore vaccination is possible upon entry to each modeled age group. The target age group can be varied to represent changing BCG vaccination policy. The vaccinated population is then modeled explicitly throughout all disease compartments. The primary action of the BCG vaccine is to prevent the transition from latent to active disease, this is included for both high and low risk latent cases. Waning vaccination effectiveness has been included by stratifying vaccine effectiveness by age group. The partial protection offered by BCG vaccination against initial infection has been included as a negative modifier on the protection from latent to active disease and as a modifier on the proportion of cases that are initially infected. This allows estimates of the effectiveness of BCG vaccination at preventing active TB disease in the susceptible population to be used, as these estimates have the most robust data sources. It is assumed that latently infected individuals do not gain additional protection from re-infection from the BCG vaccine. The BCG vaccine has been modeled as being partially protective for all individuals rather than as a "take" vaccine (i.e all or nothing protection). This assumption simplifies the model and will not impact the dynamics of TB transmission, assuming that protected and unprotected BCG vaccinated individuals obey the mass action assumption (See Chapter \@ref(introduction) and [@Anderson1991; @Keeling2007]).

@@

\begin{equation}
\lambda^k_a = \frac{\beta_a}{N} \sum\limits_{i=1}^{A} \rho_i C_{ai}\left(\frac{M\iota_i}{\nu_a^P} + \sum\limits_{j = u,v}P^j_i\right)
  (\#eq:force-of-infection)
\end{equation}

Where $\iota_a$ is the age stratified number of non-UK born pulmonary cases notified in a given year, $P^{u,v}_a$ is the number of vaccinated, and unvaccinated, pulmonary TB cases, $\rho_a$ is the age-specific proportion of cases that are smear positive, $\nu^P_a$ is the age-specific rate of starting treatment for active pulmonary TB, $C_{ai}$ is the age-stratified contact matrix (Section \@ref(contact-matrix)), $\beta_a$ is the age-stratified transmission probability, and $M$ is the mixing rate between the UK born and non-UK born population.

@@

Parameters distributions were either estimated from the available data or assumed based on common values found in the literature. Where no comparable estimates were found in the literature a largely uninformative distribution has been used. The discussion of model parameters has been split into disease model parameters, vaccination model parameters, and demographic model parameters. The data sources used to estimate model parameters have been detailed although for the Enhanced TB Surveillance system (ETS) and the Labour Force Survey more detail is provided elsewhere (Chapter \@ref(data)).
